---
title: "Batch Effect Removal with GNM"
author: "Elyas Heidari"
output:
  BiocStyle::html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(data.table)
library(tidyverse)
library(matrixcalc)
library(ComplexHeatmap)
library(visNetwork)
library(prettyR)
library(igraph)
library(destiny)
library(Matrix)
library(Rtsne)
library(ggrepel)
library(smallvis)
```


# Data
```{r pressure, message=FALSE, warning=FALSE, paged.print=FALSE}
df1 <- read.table("../data/GSE81076_D2_3_7_10_17.txt")
df2 <- read.csv("../data/GSE85241_cellsystems_dataset_4donors_updated.csv", sep = "\t")
df <- cbind(df1[intersect(rownames(df1), rownames(df2)),], df2[intersect(rownames(df1), rownames(df2)),])
# setwd("/Volumes/elihei/Internship/projects/GNK/ido/codes/")
# ido <- read.table("../data/GSE72857_umitab.txt")
# clusters <- read.csv("../data/MAP.csv", header = F)
# rownames(clusters) <- clusters$V1
# meta_data <- fread("../data/GSE72857_experimental_design.txt", skip = 19, header = T)
```

# Functions
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
graph.vis <-
  function(graph,
           directed = F,
           community = T,
           betweenness = T,
           plot = T) {
    plot_community <- function(graph, community_num) {
      t <- igraph::V(graph)$community == community_num
      v <-  igraph::V(graph)[t]
      layout <- graph$layout
      l <- matrix(layout[which(t),], nrow = sum(t), byrow = T)
      sub_graph <- igraph::induced.subgraph(graph, v)
      igraph::plot.igraph(
        sub_graph,
        vertex.size = 2 + 600 / (100 + length(v)),
        layout = l,
        main = paste('community', community_num, sep = " "),
        vertex.frame.color = igraph::V(sub_graph)$color,
        vertex.label.family = 'Helvetica',
        vertex.label.dist = -.5,
        vertex.label.cex = .6,
        vertex.label.font = 3,
        vertex.label.color = 'black'
      )
    }
    
    colrs <- c("#000000", "#FFFF00", "#1CE6FF", "#FF34FF", "#FF4A46", "#008941", "#006FA6", "#A30059",
               "#FFDBE5", "#7A4900", "#0000A6", "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87",
               "#5A0007", "#809693", "#FEFFE6", "#1B4400", "#4FC601", "#3B5DFF", "#4A3B53", "#FF2F80",
               "#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92", "#FF90C9", "#B903AA", "#D16100",
               "#DDEFFF", "#000035", "#7B4F4B", "#A1C299", "#300018", "#0AA6D8", "#013349", "#00846F",
               "#372101", "#FFB500", "#C2FFED", "#A079BF", "#CC0744", "#C0B9B2", "#C2FF99", "#001E09",
               "#00489C", "#6F0062", "#0CBD66", "#EEC3FF", "#456D75", "#B77B68", "#7A87A1", "#788D66",
               "#885578", "#FAD09F", "#FF8A9A", "#D157A0", "#BEC459", "#456648", "#0086ED", "#886F4C",
               
               "#34362D", "#B4A8BD", "#00A6AA", "#452C2C", "#636375", "#A3C8C9", "#FF913F", "#938A81",
               "#575329", "#00FECF", "#B05B6F", "#8CD0FF", "#3B9700", "#04F757", "#C8A1A1", "#1E6E00",
               "#7900D7", "#A77500", "#6367A9", "#A05837", "#6B002C", "#772600", "#D790FF", "#9B9700",
               "#549E79", "#FFF69F", "#201625", "#72418F", "#BC23FF", "#99ADC0", "#3A2465", "#922329",
               "#5B4534", "#FDE8DC", "#404E55", "#0089A3", "#CB7E98", "#A4E804", "#324E72", "#6A3A4C",
               "#83AB58", "#001C1E", "#D1F7CE", "#004B28", "#C8D0F6", "#A3A489", "#806C66", "#222800",
               "#BF5650", "#E83000", "#66796D", "#DA007C", "#FF1A59", "#8ADBB4", "#1E0200", "#5B4E51",
               "#C895C5", "#320033", "#FF6832", "#66E1D3", "#CFCDAC", "#D0AC94", "#7ED379", "#012C58",
               
               "#7A7BFF", "#D68E01", "#353339", "#78AFA1", "#FEB2C6", "#75797C", "#837393", "#943A4D",
               "#B5F4FF", "#D2DCD5", "#9556BD", "#6A714A", "#001325", "#02525F", "#0AA3F7", "#E98176",
               "#DBD5DD", "#5EBCD1", "#3D4F44", "#7E6405", "#02684E", "#962B75", "#8D8546", "#9695C5",
               "#E773CE", "#D86A78", "#3E89BE", "#CA834E", "#518A87", "#5B113C", "#55813B", "#E704C4",
               "#00005F", "#A97399", "#4B8160", "#59738A", "#FF5DA7", "#F7C9BF", "#643127", "#513A01",
               "#6B94AA", "#51A058", "#A45B02", "#1D1702", "#E20027", "#E7AB63", "#4C6001", "#9C6966",
               "#64547B", "#97979E", "#006A66", "#391406", "#F4D749", "#0045D2", "#006C31", "#DDB6D0",
               "#7C6571", "#9FB2A4", "#00D891", "#15A08A", "#BC65E9", "#FFFFFE", "#C6DC99", "#203B3C",
               
               "#671190", "#6B3A64", "#F5E1FF", "#FFA0F2", "#CCAA35", "#374527", "#8BB400", "#797868",
               "#C6005A", "#3B000A", "#C86240", "#29607C", "#402334", "#7D5A44", "#CCB87C", "#B88183",
               "#AA5199", "#B5D6C3", "#A38469", "#9F94F0", "#A74571", "#B894A6", "#71BB8C", "#00B433",
               "#789EC9", "#6D80BA", "#953F00", "#5EFF03", "#E4FFFC", "#1BE177", "#BCB1E5", "#76912F",
               "#003109", "#0060CD", "#D20096", "#895563", "#29201D", "#5B3213", "#A76F42", "#89412E",
               "#1A3A2A", "#494B5A", "#A88C85", "#F4ABAA", "#A3F3AB", "#00C6C8", "#EA8B66", "#958A9F",
               "#BDC9D2", "#9FA064", "#BE4700", "#658188", "#83A485", "#453C23", "#47675D", "#3A3F00",
               "#061203", "#DFFB71", "#868E7E", "#98D058", "#6C8F7D", "#D7BFC2", "#3C3E6E", "#D83D66",
               
               "#2F5D9B", "#6C5E46", "#D25B88", "#5B656C", "#00B57F", "#545C46", "#866097", "#365D25",
               "#252F99", "#00CCFF", "#674E60", "#FC009C", "#92896B")
    
    
    ig <- methods::as(graph, "igraph")
    
    if (betweenness) {
      bt <-
        sort(igraph::betweenness(ig, igraph::V(ig), directed = directed),
             decreasing = T)
    } else {
      bt <- NULL
    }
    community_n <- 1
    if (community) {
      fc <- igraph::cluster_louvain(igraph::as.undirected(ig))
      igraph::V(ig)$community <- igraph::membership(fc)
      community_n <- length(unique(igraph::V(ig)$community))
      mark_list <-
        lapply(c(1:community_n), function(x)
          igraph::V(ig)[igraph::V(ig)$community == x])
    } else {
      mark_list <- NULL
    }
    if (community_n <12)
      colrs <- RColorBrewer::brewer.pal(community_n + 1, "Set3")
    else {
      if (community_n + 1 <= length(colrs))
        colrs <- sample(colrs, community_n + 1)
      else
        colrs <- sample(colrs, community_n + 1, replace = T)
    }
    igraph::V(ig)$color <- colrs[igraph::V(ig)$community]
    igraph::V(ig)$label_color <- colrs[community_n + 1]
    ig$layout <- igraph::layout_nicely(ig)
    data <- visNetwork::toVisNetworkData(ig)
    if (community)
      data$nodes$group <- igraph::V(ig)$community
    vs <-
      visNetwork::visNetwork(nodes = data$nodes, edges = data$edges)  %>%
      visNetwork::visOptions(highlightNearest = list(
        enabled = TRUE,
        degree = 1,
        hover = TRUE
      ))
    if (directed)
      vs <- vs %>% visNetwork::visEdges(arrows = "to")
    
    if (plot)
      igraph::plot.igraph(
        ig,
        vertex.label = "",
        layout = ig$layout,
        vertex.size = 2 + 600 / (100 + length(igraph::V(ig))),
        vertex.color = igraph::V(ig)$color,
        vertex.frame.color = igraph::V(ig)$color
      )
    if (community) {
      if (plot) {
        for (i in 1:community_n)
          plot_community(ig, i)
      }
      com <- igraph::V(ig)$community
      names(com) <- names(igraph::V(ig))
      list(
        graph = ig,
        betweenness = bt,
        network = vs,
        communities = com
      )
    } else {
      list(graph = ig,
           betweenness = bt,
           network = vs)
    }
    
  }



ggm1 <- 
  function(data,
           methods = c("glasso"),
           community = TRUE,
           plot = FALSE,
           ...) {
    arguments <- list(...)
    threshold <- arguments$threshold
    significance <- arguments$significance
    rho <- arguments$rho

    if (is.null(threshold))
      threshold <- 0.05

    if (is.null(significance))
      significance <- 0.05

    if (is.null(rho))
      rho = 0.1

    model <- gRim::cmod( ~ . ^ ., data = data)
    S <- stats::cov.wt (data, method = "ML")$cov
    PC <- gRbase::cov2pcor(S)
    othermodels <- list()
    if ("aic" %in% tolower(methods)) {
      othermodels$aic <- aic <- gRbase::stepwise(model)
    }
    if ("bic" %in% tolower(methods)) {
      othermodels$bic <- gRbase::stepwise(model, k = log(nrow(data)))
    }
    if ("test" %in% tolower(methods)) {
      othermodels$test <- gRbase::stepwise(model, criterion = "test")
    }
    if ("threshold" %in% tolower(methods)) {
      Z <- abs(PC)
      Z[Z < threshold] <- 0
      diag(Z) <- 0
      Z[Z > 0] <- 1
      g.thresh <-  methods::as(Z, "graphNEL")
      thresh <- gRim::cmod(g.thresh, data = data)
    }
    if ("sin" %in% tolower(methods)) {
      psin <- SIN::sinUG(S, n = nrow(data))
      othermodels$gsin <-
        methods::as(SIN::getgraph(psin, significance), "graphNEL")
    }
    if ("glasso" %in% tolower(methods)) {
      C <- stats::cov2cor(S)
      res.lasso <- glasso::glasso(C, rho = rho)
      AM <- abs(res.lasso$wi) > threshold
      diag(AM) <- F
      g.lasso <- methods::as(AM, "graphNEL")
      graph::nodes(g.lasso) <- colnames(data)
      othermodels$glasso <- g.lasso
    }
    othermodels <- othermodels %>% purrr::map(methods::as, "igraph")
    commonedges <- do.call(igraph::intersection, othermodels)
    list(graph = graph.vis(
      commonedges,
      community = community,
      plot = plot,
      betweenness = T,
      directed = F
    ), wi = abs(res.lasso$wi))
  }

dist.matrix <- function(df, m=1, k){
  reflect <- function(m) {
  m[upper.tri(m)] <- t(m)[upper.tri(m)]
  m
}
  trans <- matrix.power(t.matrix, k)
  dist.pair <- function(vec1, vec2){
    sqrt(sum((vec1 %*% trans - vec2 %*% trans) ^ 2))
  }
  # to.ret <- 1:dim(df)[2] %>% map(function(x) 1:dim(df)[2] %>% map(function(y) dist.pair(df[,x], df[,y])))
  # mat <-  matrix(data=NA, nrow=dim(df)[2], ncol=dim(df)[2])
  mat <- c()
  mat <- foreach(i = 1:dim(df)[2]) %dopar% {
    # print(i)
    for(j in 1:i){
      # print(i)
      mat[j] <<- dist.pair(df[,i], df[,j])
    }
    c(mat, rep(NA, dim(df)[2] - i))
  }
  mat <- as.data.frame(do.call(rbind, mat))
  # mat <- mat %>% map(as.vector)
  # mat <- Reduce(function(x,y) pmin(x, y, na.rm = TRUE), mat)
  # mat <- matrix(mat, nrow = dim(df)[2], ncol = dim(df)[2])
  mat <- reflect(mat)
  # print(m)
  # mat <- reflect(m)
  diag(mat) <- 0
  mat
}

hv.genes <- function(df, n = 100){
  sds <- apply(df, 1, sd)
  means <- apply(df, 1, mean)
  sds <- unlist(sds)
  means <- unlist(means)
  lm <- lm(sds~means)
  sm <- summary(lm)
  res <- residuals(lm)
  names(res) <- 1:length(res)
  down <- tail(order(res), n)
  down
}


graph.col <- function(graph, lay, grp, sz){
  igraph::V(graph)$color <- rgb(0, 0, floor(255*(grp^4)/max((grp^4))), maxColorValue=255, alpha=255)
  v <-  igraph::V(graph)
  graph <- as.undirected(graph)
  plot.igraph(
    graph,
    vertex.size = sz,
    layout = lay,
    vertex.frame.color = igraph::V(graph)$color,
        vertex.label.family = 'Helvetica',
        vertex.label.dist = -.5,
        vertex.label.cex = .6,
        vertex.label.font = .3,
        vertex.label.color = 'red'
  )
}



dim.reduce <- function(data, method = "pca", annot1 = NULL, annot1.name = "annot1", annot2 = NULL, annot2.name = "annot2"){
  tsneRun <- function(distMat,perplexity=10,theta=0,max_iter=5000) {
    tsneRes <- Rtsne(distMat, perplexity = perplexity, theta = theta, 
                     max_iter = max_iter, is_distance = TRUE, dims =2)
    tsneRes <- tsneRes$Y
    rownames(tsneRes) <- colnames(distMat)
    colnames(tsneRes) <- c("x","y")
    tsneRes
  }
  if(is.null(annot1)){
    if(method == "pca"){
      pcRes <- prcomp(data, scale. = FALSE, center = TRUE)
      eigs <- pcRes$sdev^2
      eigs <- eigs/sum(eigs)
      plotTab <- data.frame(pcRes$x[,c(1,2)])
      to.ret <- ggplot(plotTab, aes(x=PC1, y = PC2)) +
        geom_point() + 
        ggtitle(paste("Scatter plot for PCA")) + 
        theme_Publication() + scale_fill_Publication() +
        labs(x = "component 1", y = "component 2")
    }
    if(method == "tsne"){
      distViab <- dist(data)
      plotTab <- data.frame(tsneRun(distViab, perplexity = 2, max_iter = 10000))
      plotTab$sampleID <- rownames(plotTab)
      to.ret <- ggplot(plotTab, aes(x=x, y = y)) +
        geom_point()+ 
        ggtitle(paste("Scatter plot for t-SNE")) +
        theme_Publication() + scale_fill_Publication() +
        labs(x = "component 1", y = "component 2")
      
    }
    if(method == "umap"){
      distViab <- dist(data)
      plotTab <- data.frame(tsneRun(distViab, perplexity = 2, max_iter = 10000))
      plotTab$sampleID <- rownames(plotTab)
      to.ret<- ggplot(plotTab, aes(x=x, y = y)) +
        geom_point() + 
        ggtitle(paste("Scatter plot for UMAP")) + theme_bw() +
        labs(x = "component 1", y = "component 2")
    }
  }else{
    if(is.null(annot2)){
      if(method == "pca"){
        pcRes <- prcomp(data, scale. = FALSE, center = TRUE)
        eigs <- pcRes$sdev^2
        eigs <- eigs/sum(eigs)
        plotTab <- data.frame(pcRes$x[,c(1,2)])
        to.ret <- ggplot(plotTab, aes(x=PC1, y = PC2, col = annot1)) +
          geom_point() + 
          ggtitle(paste("Scatter plot for PCA")) + 
          theme_Publication() + scale_fill_Publication() +
          labs(x = "component 1", y = "component 2", colour = annot1.name)
      }
      if(method == "tsne"){
        distViab <- data
        plotTab <- data.frame(tsneRun(distViab, perplexity = 2, max_iter = 10000))
        plotTab$sampleID <- rownames(plotTab)
        to.ret <- ggplot(plotTab, aes(x=x, y = y, col = annot1)) +
          geom_point()+ 
          ggtitle(paste("Scatter plot for t-SNE")) +
          labs(x = "component 1", y = "component 2", colour = annot1.name)
        
      }
      if(method == "umap"){
        distViab <- data
        plotTab <- data.frame(tsneRun(distViab, perplexity = 2, max_iter = 10000))
        plotTab$sampleID <- rownames(plotTab)
        to.ret<- ggplot(plotTab, aes(x=x, y = y, col = annot1)) +
          geom_point() + 
          ggtitle(paste("Scatter plot for UMAP")) + theme_bw() +
          labs(x = "component 1", y = "component 2", colour = annot1.name)
      }
    }else{
      if(method == "pca"){
        pcRes <- prcomp(data, scale. = FALSE, center = TRUE)
        eigs <- pcRes$sdev^2
        eigs <- eigs/sum(eigs)
        plotTab <- data.frame(pcRes$x[,c(1,2)])
        to.ret <- ggplot(plotTab, aes(x=PC1, y = PC2, col = annot1, shape = annot2)) +
          geom_point() + 
          ggtitle(paste("Scatter plot for PCA")) + 
          theme_Publication() + scale_fill_Publication() +
          labs(x = "component 1", y = "component 2", colour = annot1.name, shape = annot2.name)
      }
      if(method == "tsne"){
        distViab <- dist(data)
        plotTab <- data.frame(tsneRun(distViab, perplexity = 2, max_iter = 10000))
        to.ret <- ggplot(plotTab, aes(x=x, y = y, col = annot1, shape = annot2)) +
          geom_point()+ 
          ggtitle(paste("Scatter plot for t-SNE")) +
          theme_Publication() + scale_fill_Publication() +
          labs(x = "component 1", y = "component 2", colour = annot1.name, shape = annot2.name)
        
      }
      if(method == "umap"){
        distViab <- dist(data)
        plotTab <- smallvis(t(viabMat), method = "umap", perplexity = 40, 
                            eta = 0.01, epoch_callback = FALSE, verbose = FALSE)
        to.ret<- ggplot(plotTab, aes(x=x, y = y, col = annot1, shape = annot2)) +
          geom_point() + 
          ggtitle(paste("Scatter plot for UMAP")) +
          theme_Publication() + scale_fill_Publication() +
          labs(x = "component 1", y = "component 2", colour = annot1.name, shape = annot2.name)
      }
    }
  }
  to.ret
}





```


# GGM
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
hvg_ <- hv.genes(df, 300)
hvg <- hvg_
# df <- df[hvg, ]

# d.matrix <- dist(t(hpf[, c(hpf_sp[1:400], hpf_nc[1:400], hpf_pa[1:400], hpf_sm[1:400])]), method = "euclidean")
# write.csv(as.matrix(d.matrix), "../results/euc_all.csv")

g <- ggm1(data = data.frame(t(df[hvg,])), rho = 0.3)
# g$graph$network
l <- g$graph$graph
t.matrix <- g$wi
t.matrix <- scale(t.matrix, center=FALSE, scale=colSums(t.matrix))
# t.matrix <- g$wi
# t.matrix <- scale(t.matrix, center=FALSE, scale=colSums(t.matrix))


# hvg <- rownames(exprMat)
# g <- ggm1(data = data.frame(t(exprMat)), rho = 0.4, t.matrix = 0.05)
# g$graph$network
# t.matrix <- abs(g$wi)
# t.matrix <- ifelse(t.matrix > 0.01, t.matrix, 0)
# diag(t.matrix) <- 0
# t.matrix <- data.frame(t.matrix)
# outs <- which(colSums(t.matrix, na.rm = T) == 0)
# if(length(outs) > 0){
#   t.matrix <- t.matrix[-outs, -outs]
#   hvg <- hvg[-outs]
# }
# t.matrix <- t(scale(t.matrix, center=FALSE, scale=colSums(t.matrix)))
# CMP <- which(grepl("^CMP", meta_data$Batch_desc))
```


# GRN Metric
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# d.matrix <- dist(t(ido[,CMP]))
# write.csv(as.matrix(d.matrix), "../results/euc.csv")
# d.matrix <- read.csv("../results/euc.csv", row.names = 1)
d.matrix <- dist(t(df[hvg,]))
d.matrix_ggm <- dist.matrix(df[hvg,], k = 10)
write.csv(as.matrix(d.matrix), "../result/euc.csv")
write.csv(as.matrix(d.matrix_ggm), "../result/ggm_0.3.csv") 
d.matrix <- read.csv("../result/euc.csv", row.names = 1)
d.matrix_ggm <- read.csv("../result/ggm_0.3.csv", row.names = 1)
# cell_types[which(grepl("^MPP", cell_types))] <- "MPP"
# cell_types <- guo$num_cells

dm <- DiffusionMap(data = data.frame(sample = colnames(df), col = c(rep("batch1",dim(df1)[2]), rep("batch2",dim(df2)[2]))), distance = as.dist(d.matrix))
dpt <- DPT(dm)

# plot.DPT(dpt, dcs = c(1, 2,3), col = factor(meta_data$Batch_desc[CMP]))
df3 <- data.frame(DPT1 = dpt$DC1, DPT2 = dpt$DC2, col = dpt$col)
ggplot(df3, aes(x = DPT1, y = DPT2, color = col)) + geom_point() + theme_bw()
plot.DPT(dpt, dcs = c(1, 2))

df3 <- data.frame(DPT2 = dpt$DC2, DPT3 = dpt$DC3, col = dpt$col)
df3 <- df3[which(df3$DPT2 < .1 & df3$DPT3 < .2),]
ggplot(df3, aes(x = DPT2, y = DPT3, color = col)) + geom_point() + theme_bw()
plot.DPT(dpt, dcs = c(2, 3))

df3 <- data.frame(DPT1 = dpt$DC1, DPT3 = dpt$DC3, col = dpt$col)
df3 <- df3[which(df3$DPT1 < .1 & df3$DPT3 < .2),]
ggplot(df3, aes(x = DPT1, y = DPT3, color = col)) + geom_point() + theme_bw()
plot.DPT(dpt, dcs = c(1, 3))

tsne.plot <- Rtsne::Rtsne(d.matrix_ggm, dims = 2, perplexity = 30, pca = TRUE, max_iter = 10000, is_distance = T)
tsne.plot_euc <- Rtsne::Rtsne(d.matrix, dims = 2, perplexity = 30, pca = TRUE, max_iter = 10000, is_distance = T)
# dim.reduce(data = d.matrix, method = "tsne", annot1 = factor(meta_data$Batch_desc[CMP]), annot1.name = "cell type")
```


# Euclidean Metric
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
d.matrix <- read.csv("../results/euc.csv", row.names = 1)

dm <- DiffusionMap(data = data.frame(sample = rownames(ido)[CMP], col = factor(meta_data$Batch_desc[CMP])), distance = as.dist(d.matrix))
dpt <- DPT(dm)

df <- data.frame(DPT1 = dpt$DC1, DPT2 = dpt$DC2, col = dpt$col)
ggplot(df, aes(x = DPT1, y = DPT2, color = col)) + geom_point() + theme_bw()
plot.DPT(dpt, dcs = c(1, 2))

df <- data.frame(DPT2 = dpt$DC2, DPT3 = dpt$DC3, col = dpt$col)
ggplot(df, aes(x = DPT2, y = DPT3, color = col)) + geom_point() + theme_bw()
plot.DPT(dpt, dcs = c(2, 3))

df <- data.frame(DPT1 = dpt$DC1, DPT3 = dpt$DC3, col = dpt$col)
ggplot(df, aes(x = DPT1, y = DPT3, color = col)) + geom_point() + theme_bw()
plot.DPT(dpt, dcs = c(1, 3))

# dim.reduce(data = d.matrix, method = "tsne", annot1 = factor(meta_data$Batch_desc[CMP]), annot1.name = "cell type")
