---
title: "Lab Meeting"
author: "Elyas Heidari"
date: "7/17/2018"
output:
  BiocStyle::html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, message=FALSE, warning=FALSE, paged.print=FALSE}
library(class)
library(gmodels)
library(dbscan)
library(reshape2)
library(igraph)
library(purrr)
library(visNetwork)
library(DESeq2)
library(tximport)
library(biomaRt)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(readxl)
library(Rtsne)
library(pheatmap)
library(IHW)
library(grid)
library(gridExtra)
library(jyluMisc)
library(gtable)
library(dplyr)
library(purrr)
library(reshape)
library(reshape2)
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(highcharter)
library(dplyr)
library(purrr)
library(purrr)
library(visNetwork)
library(gRim)
library(RBGL)
library(SIN)
library(glasso)
library(igraph)
library(gRapHD)
library(bnlearn)
library(purrr)
library(visNetwork)
library(gRim)
library(RBGL)
library(SIN)
library(glasso)
library(igraph)
library(gRapHD)
library(bnlearn)
library(readxl)
library(highcharter)
library(dplyr)
library(limma)
library(ggplot2)
library(entropy)
library(purrr)
library(permute)
library(entropy)
library(purrr)
library(permute)
library(smallvis)
library(purrr)
library(visNetwork)
library(gRim)
library(RBGL)
library(SIN)
library(glasso)
library(igraph)
library(gRapHD)
library(kableExtra)

ggm <-
  function(data,
           threshold = 0.05,
           significance = 0.05,
           methods = c("glasso"),
           rho = 0.1,
           community = TRUE) {
    model <- cmod(~ . ^ ., data = data)
    S <- cov.wt (data, method = "ML")$cov
    PC <- cov2pcor(S)
    othermodels <- list()
    if("aic" %in% methods){
      othermodels$aic <- aic <- stepwise(model)
    }
    if("bic" %in% methods){
      othermodels$bic <- stepwise(model, k = log(nrow(data)))
    }
    if("test" %in% methods){
      othermodels$test <- stepwise(model, criterion = "test")
    }
    if("threshold" %in% methods){
      Z <- abs(PC)
      Z[Z < threshold] <- 0
      diag(Z) <- 0
      Z[Z > 0] <- 1
      g.thresh <- as(Z, "graphNEL")
      thresh <- cmod(g.thresh, data = data)
    }
    if("sin" %in% methods){
      psin <- sinUG(S, n = nrow(data))
      othermodels$gsin <- as(getgraph(psin, significance), "graphNEL")
    }
    if("glasso" %in% methods){
      C<-cov2cor(S)
      res.lasso <- glasso(C, rho = rho)
      AM <- abs(res.lasso$wi) > threshold
      diag(AM) <- F
      g.lasso <- as(AM, "graphNEL")
      nodes(g.lasso) <- colnames(data)
      othermodels$glasso <- g.lasso
    }
    othermodels <- othermodels %>% map(as, "igraph")
    commonedges <- do.call(igraph::intersection, othermodels)
    bt <- betweenness(as(commonedges, "igraph"), V(as(commonedges, "igraph")))
    data <- toVisNetworkData(as(commonedges, "igraph"))
    if(community){
      fc <- cluster_louvain(as(commonedges, "igraph"))
      data$nodes$group <- membership(fc)
    }
    vs <- visNetwork(nodes = data$nodes, edges = data$edges)  %>%
      visOptions(highlightNearest = list(
        enabled = TRUE,
        degree = 1,
        hover = TRUE
      ))
    list(graph = as(commonedges, "igraph"), betweenness = bt, network = vs)
  }
```



#P1: 3'RNASeq data 
Analysis of 3'RNASeq data of 15 cell lines and sgTP53 knock-down treatment.

## Preprocess
```{r ensembl, message=FALSE, warning=FALSE, paged.print=FALSE}
load("../../3RNAseq_180705_/data/RData/3RNAseq.RData")
sample.list <- read_xlsx("../../3RNAseq_180705_/data/metadata/sampleList.xlsx", sheet = 2)
plate <- sample.list$`Plate well number`
id <- sample.list$ID
sample.name<- sample.list$`Sample Name`
cell_line <- sample.name %>% map(function(x) ifelse(length(strsplit(x, split = "_")[[1]]) == 4, paste(strsplit(x, split = "_")[[1]][c(2,3)], collapse = "_"), strsplit(x, split = "_")[[1]][2]))
cell_line <- unlist(cell_line)
treatment <- sample.name %>% map(function(x) ifelse(length(strsplit(x, split = "_")[[1]]) == 4, paste(strsplit(x, split = "_")[[1]][4], collapse = "_"), strsplit(x, split = "_")[[1]][3]))
treatment <- unlist(treatment)
sample.annotations <- data.frame(row.names = id, cell_line = as.character(cell_line), treatment = as.character(treatment), plate = plate)
# write.csv(sample.annotations, file = "../data/metadata/sample_annotations.csv", row.names = T)
colData(dds) <- cbind(colData(dds), sample.annotations[rownames(colData(dds)),])
dds <- estimateSizeFactors(dds)
dds <- dds[which(rowSums(counts(dds)) > 50),]
dds.vst <- varianceStabilizingTransformation(dds, blind = TRUE)
exprMat <- assay(dds.vst)
sds <- rowSds(exprMat)
exprMat <- exprMat[order(sds, decreasing = TRUE)[1:5000],]
```

## PCA on Expression Matirix
```{r PCA, message=FALSE, warning=FALSE, paged.print=FALSE}
pcRes <- prcomp(t(exprMat), scale. = FALSE, center = TRUE)
eigs <- pcRes$sdev^2
eigs <- eigs/sum(eigs)

plotTab <- data.frame(pcRes$x[,c(1,2)])
plotTab$sampleID <- rownames(plotTab)
plotTab$cellLine <- sample.annotations[plotTab$sampleID, "cell_line"]
plotTab$treatment <- sample.annotations[plotTab$sampleID, "treatment"]

ggplot(plotTab, aes(x=PC1, y = PC2, label = sampleID, color = cellLine, shape = treatment)) +
  geom_point() +
  geom_text_repel(aes(PC1,PC2, label = sampleID)) +
  theme_bw()
```

## t-SNE on Expression Matirix
```{r t-SNE, message=FALSE, warning=FALSE, paged.print=FALSE}
tsneRun <- function(distMat,perplexity=10,theta=0,max_iter=5000) {
  tsneRes <- Rtsne(distMat, perplexity = perplexity, theta = theta, 
                   max_iter = max_iter, is_distance = TRUE, dims =2)
  tsneRes <- tsneRes$Y
  rownames(tsneRes) <- labels(distMat)
  colnames(tsneRes) <- c("x","y")
  tsneRes
}

distViab <- dist(t(exprMat))

plotTab <- data.frame(tsneRun(distViab, perplexity = 2, max_iter = 10000))

#plot
plotTab$sampleID <- rownames(plotTab)
plotTab$cellLine <- sample.annotations[plotTab$sampleID, "cell_line"]
plotTab$treatment <- sample.annotations[plotTab$sampleID, "treatment"]

ggplot(plotTab, aes(x=x, y = y, label = sampleID, color = cellLine, shape = treatment)) +
  geom_point() +
  geom_text_repel(aes(x,y, label = sampleID)) +
  theme_bw()
```

## GGM on Expression Matirix
```{r GGM, message=FALSE, warning=FALSE, paged.print=FALSE}
dds <- estimateSizeFactors(dds)
dds <- dds[which(rowSums(counts(dds)) > 50),]
dds.vst <- varianceStabilizingTransformation(dds, blind = TRUE)
exprMat <- assay(dds.vst)
sds <- rowSds(exprMat)
exprMat <- exprMat[order(sds, decreasing = TRUE)[1:5000],]
g <- ggm(data.frame(exprMat), rho = 0.5)
g$network
```

Results from GGM and t-SNE are exactly the same. Maybe time for paradigm shit! :-?!


#P2: Connected KNN graph
KNN graphs may be not connected but we are intrested in conncected graphs! They are important in single cell analysis in high-dimentions.

## KNN graph
```{r knn_graph, message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr)
graph.knn <- function(nn) {
  m <- melt(nn$id)
  d <- melt(nn$dist)
  m <- m %>% dplyr::select(X1, value)
  g <- graph_from_data_frame(m, directed = FALSE)
  d <- d %>% dplyr::select(X1, value)
  g <- g %>% set_edge_attr("weight", value = d$value)
  g
}

is.con <- function(data, k) {
  nn <- kNN(data, k = k)
  g <- graph.knn(nn)
  is.con <- list()
  is.connected(g, mode = c("weak", "strong"))
}


add.edges <- function(data, k.min) {
  k <- k.min
  while (!is.con(data, k)) {
    k <- k + 1
  }
  nn.min <- kNN(data, k = k.min)
  g.min <- graph.knn(nn.min)
  clus <- clusters(g.min)
  clus.m <- clus$membership
  nn.max <- kNN(data, k = k)
  g.max <- graph.knn(nn.max)
  g.dif <- difference(g.max, g.min)
  edges <- as.data.frame(get.edgelist(g.dif))
  edges$V4 <- E(g.dif)$weight
  while (!is.connected(g.min, mode = c("weak", "strong"))) {
    cl <- clusters(g.min)
    cl.m <- cl$membership
    edges %>% dplyr::mutate(V3 = (cl.m[V1] != cl.m[V2])) -> edges
    edges %>% dplyr::filter(V3 == TRUE) -> edges
    to.add1 <- edges$V1[which.max(edges$V4)]
    to.add2 <- edges$V2[which.max(edges$V4)]
    g.min <- add_edges(g.min, c(to.add1, to.add2))
  }
  g.min <- igraph::simplify(g.min, remove.multiple = TRUE, remove.loops = TRUE)
  data <- toVisNetworkData(g.min)
  nodes <- data$nodes
  edges <- data$edges
  groups <- clus.m[nodes$id]
  groups -> nodes$group
  vs <- visNetwork(nodes = nodes, edges = edges)  %>%
      visOptions(highlightNearest = list(
        enabled = T,
        degree = 1,
        hover = T
      )) 
  vs
}


# add.edges(iris[,1:4], 3)
# data <- read.csv("/Volumes/C/2018/Questools/data/data.csv")
# add.edges(data[1:1000,], 3)

add.edges(iris[,1:4], 3)
data <- read.csv("/Volumes/C/2018/Questools/data/data.csv")
# add.edges(data[1:1000,], 3)
add.edges(t(data[,-1]), 3)
data <- read.csv("/Volumes/elihei/Internship/3RNAseq_180705_/data/aux_data/exprMat.csv")
dim(data)
system.time(add.edges(t(data[,-1]), 3))
```


#P3: TP53ness

Use machine learning and statistical apoaches on multi-omics data to define and detect TP53ness. We have found something already!

```{r TP53ness, message=FALSE, warning=FALSE, paged.print=FALSE}

################### data 
load("../../TP53ness_180716_/data/metadata/patmeta_180504.RData")
load("../../TP53ness_180716_/data/RNASeq/ddsrna_180301.RData")
load("../../TP53ness_180716_/data/Metabolism/Seahorse_20170822_combat.RData")
load("../../TP53ness_180716_/data/FISH/fish_20170604.RData")
screen.files <- list.files("../../TP53ness_180716_/data/screen/", pattern = ".*\\.rds")
screens <- screen.files %>% map(function(f) readRDS(paste0("../../TP53ness_180716_/data/screen/", f)))
names(screens) <- unlist(screen.files %>% map(function(x) gsub(x, pattern = "\\.rds", replacement = ""))) 
wes <- read.csv("../../TP53ness_180716_/data/WES/2018-04-09_wes_formatted.csv")
ampliSeq <- read.csv("../../TP53ness_180716_/data/ampliSeq/ampliSeq_180509.csv")
ighv <- read.csv("../../TP53ness_180716_/data/IGHV/2018-03-05_IGHV_combined.csv")
wgs <- read.csv("../../TP53ness_180716_/data/WGS/2018-04-13_wgs_formatted.csv")


################### preprocess
patTP53 <- patMeta[which(!is.na(patMeta$TP53)),]
rownames(patTP53) <- patTP53$Patient.ID 


################### RNASeq
# dds <- estimateSizeFactors(dds)
# dds <- dds[apply(counts(dds), 1, function(x) any(x > 10)),]
# dds.vst <- varianceStabilizingTransformation(dds, blind = TRUE)
# exprMat <- assay(dds.vst)
# sds <- rowSds(exprMat)
# names(sds) <- rownames(exprMat)
# exprMat <- exprMat[names(sort(sds, decreasing = TRUE)[1:5000]),]
# exprMat <- data.frame(exprMat)
# write.csv(exprMat, "../exprMat.csv")
exprMat <- read.csv("../exprMat.csv")
rownames(exprMat) <- exprMat$X
exprMat <- exprMat[,-1]
rna.data <- t(exprMat)

################### Seahorse
sea.data <- sea$Viability
sea.data <- data.frame(sea = sea.data, row.names = sea$patientID)

################### FISH
## very sparse!

################### CPS1000
cps.data <- screens$CPS1000
cps.data <- cps.data %>% dplyr::mutate(drug.cons = paste(Drug, Concentration, sep = "."))
cps.data <- cps.data %>% dplyr:: select(patientID, drug.cons, normVal.cor)
cps.data <- cps.data %>% dplyr::group_by(patientID, drug.cons) %>% dplyr::summarise(mean(normVal.cor))
colnames(cps.data)[3] <- "normVal.cor"
cps.data <- dcast(cps.data, patientID ~ drug.cons)
rownames(cps.data) <- cps.data$patientID
cps.data <- cps.data[,-1]

################### Binding
inr <- Reduce(intersect, list(rownames(cps.data), rownames(rna.data), patTP53$Patient.ID))
d <- Reduce(cbind, list(rna.data[inr,], cps.data[inr,]))
rownames(d) <- paste0(rownames(d), (patTP53[inr, "TP53"] == 1))
l <- ggm(data.frame(t(d)), rho = 0.3)
g <- l$graph

data1 <- toVisNetworkData(as(g, "igraph"))
group <- as.data.frame(patTP53[inr, "TP53"])$TP53
names(group) <- inr
data1$nodes$group <- group
visNetwork(nodes = data1$nodes, edges = data1$edges)  %>%
  visOptions(highlightNearest = list(
    enabled = TRUE,
    degree = 1,
    hover = TRUE
  ))
```



